---
title: "Washington Bike Sharing Service"
author: "Maurizio Murino"
date: "16 February 2016"
output: html_document
---
# 1. Introduction

_This project is about the Capital Bikeshare (CaBi) scheme (<http://www.capitalbikeshare.com/>) in Washington DC, launched in 2008 by the government of District of Columbia. In this project, we focus on the operations of CaBi in 2011 and 2012. In the Excel spreadsheet `data.xlsx`, it contains hourly rental data spanning these two years. Two types of rental data were recorded: casual
renters and registered renters. A registered renter is a pre-registered member of the bike sharing program, while a casual renter pays for the use of a bike on the spot using a credit card. In addition to the time/date information (including holidays), the spreadsheet also records the weather condition in each hour (e.g., temperature, wind speed and humidity)._  

# 2. Loading the data

We are going to load in R the whole Excel workbook. Since it is composed by three separated sheets, the package `XLConnect` comes in use. It allow to read the entire workbook, and then load the single worksheets. Please note that `XLConnect` requires Java to be pre-installed. 

```{r introduction 1, cache=TRUE, results='hide'}
library(XLConnect)
work_book <- loadWorkbook("raw data.xlsx") 
```

Since the `work_book` is quite small (1728 bytes) we are going to load in memory every sheet like a separate object.

```{r introduction 2}
object.size(work_book)
```

```{r introduction 3, cache=TRUE}
note <- readWorksheet(work_book, sheet = "Note")
all_data <- readWorksheet(work_book, sheet = "All data")
testing <- readWorksheet(work_book, sheet = "Testing")
```

# 3. Data description

The Excel workbook is composed by three sheets: "Note" is the coodebook, "All data" is the main dataset, "Testing" is a second and smaller dataset.

```{r data description 1, results='hide'}
summary(work_book)
```

### 3.1 "Note"

It is the codebook. It describes how the content of the dataset is coded. Specifically, it reports how seasons (`season`) and weekdays (`Weekday`) are numerically coded in the datasets.

```{r data description 2}
str(note)
```

###3.2 "all_data"

The dataset `all_data` is composed by 17379 observation with 13 variables. `datetime` is profitably flaked by variables providing further information on the day of the week, the season and if we are talking about a working or a resting day. 

```{r data description 3}
str(all_data)
```
